### 2016-07-06 motivation.tex
Edits to the example.

In this section, we motivate our ideas via an example written in
\txnimp - a C-like imperative language equipped with a \C{txn} lexical
block defining a transaction scope.  For expositional convenience,
each \C{txn} block is associated with a single-quoted string in angle
braces that uniquely identifies the transaction. We use Hoare triple
notation to annotate programs with pre- and post- conditions.

@@

\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
it wrote to \C{B}, and the value of \C{B} is either \C{k-a1-a2} or
\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
remains invariant until \C{Wd1} commits, then the post-condition (\C{B
= k-a1-a2}) can be established easily. However, an interference from
\C{Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{B} and committing \C{Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from \C{Wd2} starting from
an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds; $\C{Wd2} \hboar \C{Wd1}$ does not hold because
\C{Wd1}'s write to \C{B} clearly happened before \C{Wd2}'s commit, and
$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
committed, while \C{Wd2} has already begun. Since \C{Wd2} also writes
to \C{B}, this violates the $\psi_{SI}$ constraint which we assume to
be an invariant. A proof that the post-condition holds now follows
from the contradiction. It is informative to note that if the
invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
contradiction and we cannot rule out the interference, which
(rightfully) causes  the proof to fail.

>>>>

In this section, we motivate our ideas via an example written in
\txnimp - a C-like imperative language equipped with a \C{txn} lexical
block defining a transaction scope. Each \C{txn} block is associated
with an indentifier in angle braces that uniquely identifies the
transaction. We use Hoare triple notation to annotate programs with
pre- and post- conditions.

@@

\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
it wrote to \C{B}, and the value of \C{B} is either \C{k-a1-a2} or
\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
remains invariant until \C{Wd1} commits, then the post-condition (\C{B
= k-a1-a2}) can be established easily. However, an interference from
\C{Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a2}$ to \C{B} and committing \C{Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from \C{Wd2} starting from
an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds; $\C{Wd2} \hboar \C{Wd1}$ does not hold because
\C{Wd1}'s write to \C{B} clearly happened before \C{Wd2}'s commit, and
$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
committed, while \C{Wd2} has already begun. Since \C{Wd2} also writes
to \C{B}, this violates the $\psi_{SI}$ constraint which we assume to
be an invariant. A proof that the post-condition holds now follows
from the contradiction. It is informative to note that if the
invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
contradiction and we cannot rule out the interference, which
(rightfully) causes  the proof to fail.

### 2016-07-06 opsem.tex
Edits to the example.

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transaction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp\
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

@@

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
below) is not a transitive relation, and hence lets us capture
finer-grained isolation properties than $\hbZ$, which we leverage in
our development.  The session order relation captures the sequential
order of operations within a transaction. In particular, it relates
two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
monotonically increasing identifiers to effects, as defined by the
$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}. Evaluation
contexts ($\ectx_i$) for transaction-bound terms are defined so as to
enforce a deterministic sequential order of execution within a
transaction, leading to a deterministic total order among effect ids,
which defines the session order relation. Visibility ($\visZ$) on the
other hand relates effects across concurrent transactions.
Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
during its execution, thus effecting its return value
($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
the value ($\rval$) of the write effect with the highest id among its
visible effects (this is made possible by appropriately defining
$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
value of a read depends on what write effects it can witness. An
operation can only witness the effects of already concluded
operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions.

>>>>

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace ($\E$).
The first component of the trace is a set ($\A$) of \emph{effects},
where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
the transaction that generated this effect, (b). an indentifier ($j$)
that uniquely identifies this effect, (c). the operation ($\op$)
documented by this effect, which can be either a read (\C{RD(X)}), a
write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
value, if any, associated with the operation (\eg, value read or value
written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
of the four aforementioned individual components of the tuple.
In every step of the evaluation, the machine reduces a \txnimp\ term
by executing a read, write or commit operation, generating an effect,
and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

@@

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
below) is not a transitive relation, and hence lets us capture
finer-grained isolation properties than $\hbZ$, which we leverage in
our development.  The session order relation captures the sequential
order of operations within a transaction. In particular, it relates
two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
monotonically increasing identifiers to effects, as defined by the
$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux} ($\maxId(\A)$
returns the maximum number identifying an effect in $\A$). Evaluation
contexts ($\ectx_i$) for transaction-bound terms are defined so as to
enforce a deterministic sequential order of execution within a
transaction, leading to a deterministic total order among effect ids,
which defines the session order relation. Visibility ($\visZ$) on the
other hand relates effects across concurrent transactions.
Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
during its execution, thus effecting its return value
($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
the value ($\rval$) of the write effect with the highest id among its
visible effects (this is made possible by appropriately defining
$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
value of a read depends on what write effects it can witness. An
operation can only witness the effects of already concluded
operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions.

### 2016-07-06 reasoning.tex
Edits to the example.

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away environmental
interference as a rely relation. A conspicuous difference between a
standard development of rely-guarantee and ours is that, while the
former reasons in terms of program states (variable to value
bindings), we reason in terms of executions as captured by their
traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and our pre- ($P$) and post- ($Q$) conditions are assertions over
executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
also facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the post-condition
$\C{B\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{B} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{B} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{B}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(B) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(B) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}

@@

\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to
the semantics of Definition~\ref{def:rg-semantics}.\footnote{Full proofs are provided in the supplementary material.}
\end{theorem}

>>>>

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away environmental
interference as a rely relation. A conspicuous difference between a
standard development of rely-guarantee and ours is that, while the
former reasons in terms of program states (variable to value
bindings), we reason in terms of executions as captured by their
traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and our pre- ($P$) and post- ($Q$) conditions are assertions over
executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
also facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}, which interprets
the given set of effects as a state. For example, if a \emph{bi-state}
rely relation relates every pair of states $\sigma$ and $\sigma'$ such
that $\sigma'(X) \ge \sigma(X) \ge 0$, the corresponding
\emph{bi-execution} rely relation relates every pair of executions
$\E$ and $\E'$ such that $\interp{\E.\A}(X) \ge \interp{\E'.\A}(X) \ge
0$ . Assertions on states are also written similarly. For instance,
consider the post-condition $\C{B\,=\,k-a1-a2}$ of the program in
Fig.~\ref{fig:motiv-eg-1}. The corresponding assertion on the
post-state ($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts
that in all states resulting from executing the program, the value of
\C{B} is \C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E.\A}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in
all executions of the program, the value of \C{B} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{B}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E.\A}(B) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E.\A}(B) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}

@@

\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
is provided in the supplementary material.}
\end{theorem}

### 2016-07-06 store-consistency.tex
Edits to the example.

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of the
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is redefined
to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
honor its consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailishat} is often used in practice to capture the
best-effort nature of {\sc sc}.}. We generalize this approach to any
$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
determine an acceptable weakening of $\I_s$ in case of a conflict with
$\I_c$.  The principle requires the weakened consistency guarantee
($\I_s'$) of the store to enforce all visibility relationships imposed
by the actual consistency guarantee ($\I_s$), unless enforcing such a
relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
% GK : Please talk to me before changing the substance of the following paragraph.
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes {\sc sc} stores with on-demand weak isolation, such as
provided by off-the-shelf relational databases. Similar approaches can be employed
to capture the semantics of other kinds of stores, such as a causally
consistent ({\sc cc}) data store, as specific instantiations of our
operational semantics\footnote{Details of the {\sc cc} instantiation
are given in the supplementary material.}.

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. We assume an {\sc sc} store, such as a relational
database, whose store-specific trace invariant ($\I_s$) was shown
previously. Both transactions are run at {\sc si} isolation, hence
$\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
shown\footnote{The supplementary material shows the fully decorated program.}.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{C-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state.  The context for propositions is also the implicit,
i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
$\ge$ a1+a2} remains an invariant, hence elided. The pre-condition
$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
before \C{Wd1}, writing \C{k-a2} to \C{B}. Since neither \C{Wd1} nor
\C{Wd2} are committed at the beginning, the pre-condition $P$ of the
program is extended with $\neg\committed(\C{Wd1}) \wedge
\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
$Q_1$ of \C{Wd1} asserts different values for \C{B} depending on
whether or not \C{Wd2} has already committed.  It also asserts that
\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
also satisfy its spec, but the proof is similar).

@@

\noindent The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions now write to \C{B}, {\sc si} requires either
$\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si} (which
subsumes {\sc rc}) prohibits the former possibility, allowing us the
deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
${\committed(\C{Wd2})}  \Rightarrow \C{B = k-a2} \wedge \C{B =
k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.

@@

The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected,
and if it commits \C{Wd2}, then $\C{Wd2}\not\visar\C{Wd1}$, hence
$\phi_4$ doesn't care. For $\phi_5$, {\sc si} condition should be used
to show that any interference from \C{Wd2} is invalid. The proof
proceeds on the same lines as the Hoare proof discussed above; it is
also discussed in detail in \S\ref{sec:motivation}.

>>>>

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of the
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is redefined
to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
honor its consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailishat} is often used in practice to capture the
best-effort nature of {\sc sc}.}. We generalize this approach to any
$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
determine an acceptable weakening of $\I_s$ in case of a conflict with
$\I_c$.  The principle requires the weakened consistency guarantee
($\I_s'$) of the store to enforce all visibility relationships imposed
by the actual consistency guarantee ($\I_s$), unless enforcing such a
relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
\label{def:maxvis}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
% GK : Please talk to me before changing the substance of the following paragraph.
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes {\sc sc} stores with on-demand weak isolation, such as
provided by off-the-shelf relational databases. Similar approaches can be employed
to capture the semantics of other kinds of stores, such as a causally
consistent ({\sc cc}) data store, as specific instantiations of our
operational semantics\footnote{Details of the {\sc cc} instantiation
are given in the supplementary material.}.

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. Both withdraw transactions are run at {\sc si}
isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd1)}}$. We assume an {\sc sc} store, such as a
conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
the maximul visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
w.r.t. $\I_c$, as described in \S\ref{sec:store-consistency}. In
concrete terms, this means that an effect added to $\E.\A$ by the
current thread or a concurrent thread immediately becomes visible to
the subsequent operations unless we produce a witness that such
visibility violates $\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note
that the language of our specifications and (most of) the assertions
on executions is a decidable subset of the first-order logic (in
particular, our specifications and rely-guarantee rules are carefully
crafted to avoid transitive closures and quantification over infinite
domains). Consequently, the proof obligations generated by the
application of our rules can often be discharged via an SMT solver. We
nonetheless give full details in the following.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not shown.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{B-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state. The context for propositions the value computation of
shared variables is also the implicit, i.e., we write $\psi$ instead
of $\underE{\psi}$ and $\C{B = k}$ instead of $\C{\interp{\E.\A}(B) =
k}$. The proposition \C{k $\ge$ a1+a2} remains an invariant, hence
elided. The pre-condition $P_1$ of \C{Wd1} accounts for the
possibility of \C{Wd2} committing before \C{Wd1}, writing \C{k-a2} to
\C{B}. Since neither \C{Wd1} nor \C{Wd2} are committed at the
beginning, the pre-condition $P$ of the program is extended with
$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
$P_1$ follows. The post-condition $Q_1$ of \C{Wd1} asserts different
values for \C{B} depending on whether or not \C{Wd2} has already
committed.  It also asserts that \C{Wd1} has committed. The
post-condition $Q_2$ of \C{Wd2} (not shown) is similar, but with the
roles of \C{Wd1} and \C{Wd2} exchanged. The post-condition $Q$ follows
from $Q_1 \wedge Q_2$, hence valid. It remains to show that \C{Wd1}
satisfies its specification (\C{Wd2} must also satisfy its spec, but
the proof is similar).

@@

\noindent The rule also lets us assume that the execution satisfies
trace invariant ($\I$), which asserts {\sc si} for both transactions.
Since both transactions now write to \C{B}, {\sc si} requires either
$\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si}, which
subsumes {\sc rc} and requires only committed transactions be visible,
prohibits $\C{Wd1} \visar \C{Wd2}$, allowing us the deduce $\C{Wd2}
\visar \C{Wd1}$. This lets us derive ${\committed(\C{Wd2})}
\Rightarrow \C{B = k-a2} \wedge \C{B = k-a2-a1}$, allowing us to prove
the post-condition $\phi_5$.

@@

The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected.
If it commits \C{Wd2}, then since \C{Wd1} had already performed (read)
operations that cannot witness \C{Wd2}, we have
$\C{Wd2}\not\visar\C{Wd1}$, and since $\phi_4$ only makes assertions
under the case $\C{Wd2}\visar\C{Wd1}$, it is stable. For $\phi_5$,
{\sc si} condition should be used to show that any interference from
\C{Wd2} is invalid. The proof proceeds on the same lines as the Hoare
proof discussed above; it is also discussed in detail in
\S\ref{sec:motivation}.
